Conceptual and methodological problems with bias detection and avoidance in natural language processing

### 2021-06-15 thesis/02-cosine.Rmd
Adds up-to-date changes

@manzini2019black present a different approach towards finding similarities between classes of words. The authors claim that texts available online are full of direct or indirect human stereotypes. As a result, word embeddings are prone to learn and maybe amplify those biases and propagate them further into AI models that are used for various applications. Cosine distance is used in the article as a measure to first argue for the existence of multi-class bias and then to show how through bias mitigation techniques the bias may be decreased.

@@

Assuming for a moment that the value of multi-class cosine distance is correct, one may question the results' interpretation. In @manzini2019black Table 2, there are summarized the averages of cosine distance per group (gender, race, religion). I would like to focus  now on analyzing the values relating to religious biases. Here is the fragment of table that refers to that:

>>>>

@Manzini2019blackToCriminal present a different approach towards finding similarities between classes of words. The authors claim that texts available online are full of direct or indirect human stereotypes. As a result, word embeddings are prone to learn and maybe amplify those biases and propagate them further into AI models that are used for various applications. Cosine distance is used in the article as a measure to first argue for the existence of multi-class bias and then to show how through bias mitigation techniques the bias may be decreased.

@@

Assuming for a moment that the value of multi-class cosine distance is correct, one may question the results' interpretation. In @Manzini2019blackToCriminal Table 2, there are summarized the averages of cosine distance per group (gender, race, religion). I would like to focus  now on analyzing the values relating to religious biases. Here is the fragment of table that refers to that:

### 2021-06-15 thesis/03-bayesian.Rmd
Adds up-to-date changes

We also applied to the word embeddings hard debiasing method found in @manzini2019black. After debiasing the Reddit word embeddings we calculated the cosine distance again to see if there are any significant changes in terms of the similarity between protected words and the attributes.

@@

One may notice that in religion data set

It is worth paying attention to gender debiased dataset. The change is really minor, still zero out of HPDI range. Note how `different` attributes have high similarity value.

At the same time, there is a small improvement in race, at the price of moving neutral terms closer

In @manzini2019black it is mentioned that debiasing with this method that is based on bias component removal may be insufficient. The reason is that words containing similar biases may still cluster together even after debiasing.

>>>>

We also applied to the word embeddings hard debiasing method found in @Manzini2019blackToCriminal. As it was mentioned before, debiasing consists of two components: identifying the bias subspace and then removing this vector subspace from the chosen embeddings. The aim is to increase the cosine distance between protected words and the set of attributes.
After debiasing the Reddit word embeddings we calculated the cosine distance again to see if there are any significant changes in terms of the similarity between protected words and the attributes.

@@

One may notice that in a religion data set most of the distances for `associated` words slightly increased. Some of the distances from `different` class are still smaller than for the `associated` class. The distances for all classes seem in most cases to cluster together even after debiasing. The reason may be that uncertainty is quite high in `associated` and `different` class because the number of samples is so small. Additionally in some cases distances from `human` class are smaller than for the `assosiated` class which also makes it less clear how to interpret the results, if the distances are really able to catch the bias presence.

It is worth paying attention to the gender debiased dataset. The change is really minor, still zero out of HPDI range. The `different` attributes still have high similarity value. It seems that the distances for gender protected words are still clustered together which suggests that this method is not able to catch the complex bias presence. One could conclude that co-occurrence of certain terms does not always mean direct connection in terms of terms associations. The fact that some of stereotypically associated with female attributes still have high similarity with male protected words suggest that other techniques should be applied to detect and remove the bias.

At the same time, there is a small improvement in race dataset. In some cases the distances just cluster together. In general distances for `assosiated` class are still higher than those for the rest of classes. One may assume that there reason for so small change is that either the metric does not indicate properly biases or that the issue is more complex and subtracting subspaces is not enough.

In @Manzini2019blackToCriminal it is mentioned that debiasing with this method that is based on bias component removal may be insufficient. The reason is that words containing similar biases may still cluster together even after debiasing.

### 2021-06-15 thesis/04-discussion.Rmd
Adds up-to-date changes

- 89% highest posterior density intervals

>>>>

- the distributions of cosine distances of whole population and individual words can vary enormously
- including uncertainty shows how in some cases it is not clear why a word is according to this method
classified as anomalous
- debiasing with the method provided by @Manzini2019blackToCriminal does not solve the bias issues

One should remember that there is no baseline for the interpretation of the averaged cosine distances in @Manzini2019blackToCriminalToCriminal. One may assume that if the cosine distance is close to 1 then it is a desired outcome as it means, according to cosine similarity assumptions, that there is almost no similarity between the words. However what does it mean to be close to 0? If the averaged cosine similarity is equal to 80, then should we still debiase it? It is unclear what the criteria are. On one hand, it seems to be beneficial when the outcome is simplified as it is easier to compare results with one value per set. On the other hand, it is prone to misunderstanding of how to interpret the results and what threshold to assume.

We propose the use of Bayesian method to introduce the uncertainty measure in bias detection.

# Summary

The aim of this paper was to show the analysis of bias detection methods used in modern natural language processing projects. 
In the further research it is worth to dive deeper into some issues or to examine new ideas that occurred during our work.

<!-- 
remarks:
- delete papers from github/hide them
-

-->

### 2021-06-15 thesis/_book/_main.tex
Adds up-to-date changes

We first encountered the GoogleNews word embedding model in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}). They use it for their calculations as mentioned in \href{https://github.com/tolga-b/debiaswe}{Tolga Code}. We decided to compare the results obtained with the use of Reddit L2 model and the ones that we got by GoogleNews model. The details of GoogleNews model can be found here \href{https://code.google.com/archive/p/word2vec/}{Google News model}. One of the main differences between the models is that Reddit word embeggings have 50 dimensions and GoogleNews word embeddings have 300 dimensions. As the dimension is increasing, the vector can capture much more information but this is not always the case. According to some researchers the amount of dimensions should be chosen by taking into account the type of corpus and its features. In our case both Reddit and GoogleNews models are already trained and ready to use so we do not analyse further the choice of their hyperparameters.

@@

We also applied to the word embeddings hard debiasing method found in (\protect\hyperlink{ref-manzini2019black}{\textbf{manzini2019black?}}). After debiasing the Reddit word embeddings we calculated the cosine distance again to see if there are any significant changes in terms of the similarity between protected words and the attributes.

@@

One may notice

\begin{itemize}
\item
  humans in religion got closer (should they?)
\item
  change in Gender is really minor, still zero out of HPDI range, note how different is very similar
\item
  small improvement in race, at the price of moving neutral terms closer
\end{itemize}

In (\protect\hyperlink{ref-manzini2019black}{\textbf{manzini2019black?}}) it is mentioned that debiasing with this method that is based on bias component removal may be insufficient. The reason is that words containing similar biases may still cluster together even after debiasing.

@@

\begin{itemize}
\tightlist
\item
  the distributions of cosine distances of whole population and individual words can vary enormously
\item
  including uncertainty shows how in some cases it is not clear why a word is according to this method
  classified as anomalous
\item
  debiasing with the method provided by (\protect\hyperlink{ref-manzini2019black}{\textbf{manzini2019black?}}) does not solve the bias issues
\end{itemize}

@@

>>>>

We first encountered the GoogleNews word embedding model in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}). They use it for their calculations as written in \href{https://github.com/tolga-b/debiaswe}{Tolga Code}. We decided to compare the results obtained with the use of Reddit L2 model and the ones that we got by GoogleNews model. The details of the model can be found here \href{https://code.google.com/archive/p/word2vec/}{Google News model}. One of the main differences between the models is that Reddit word embeggings have 50 dimensions and GoogleNews word embeddings have 300 dimensions. As the dimension is increasing, the vector can capture much more information but this is not always the case. According to some researchers the amount of dimensions should be chosen by taking into account the type of corpus and its features. In our case both Reddit and GoogleNews models are already trained and ready to use so we do not analyse further the choice of their hyperparameters.

@@

We also applied to the word embeddings hard debiasing method found in \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}). As it was mentioned before, debiasing consists of two components: identifying the bias subspace and then removing this vector subspace from the chosen embeddings. The aim is to increase the cosine distance between protected words and the set of attributes.
After debiasing the Reddit word embeddings we calculated the cosine distance again to see if there are any significant changes in terms of the similarity between protected words and the attributes.

@@

One may notice that in a religion data set most of the distances for \texttt{associated} words slightly increased. Some of the distances from \texttt{different} class are still smaller than for the \texttt{associated} class. The distances for all classes seem in most cases to cluster together even after debiasing. The reason may be that uncertainty is quite high in \texttt{associated} and \texttt{different} class because the number of samples is so small. Additionally in some cases distances from \texttt{human} class are smaller than for the \texttt{assosiated} class which also makes it less clear how to interpret the results, if the distances are really able to catch the bias presence.

It is worth paying attention to the gender debiased dataset. The change is really minor, still zero out of HPDI range. The \texttt{different} attributes still have high similarity value. It seems that the distances for gender protected words are still clustered together which suggests that this method is not able to catch the complex bias presence. One could conclude that co-occurrence of certain terms does not always mean direct connection in terms of terms associations. The fact that some of stereotypically associated with female attributes still have high similarity with male protected words suggest that other techniques should be applied to detect and remove the bias.

At the same time, there is a small improvement in race dataset. In some cases the distances just cluster together. In general distances for \texttt{assosiated} class are still higher than those for the rest of classes. One may assume that there reason for so small change is that either the metric does not indicate properly biases or that the issue is more complex and subtracting subspaces is not enough.

In \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) it is mentioned that debiasing with this method that is based on bias component removal may be insufficient. The reason is that words containing similar biases may still cluster together even after debiasing.

@@

\begin{itemize}
\tightlist
\item
  the distributions of cosine distances of whole population and individual words can vary enormously
\item
  including uncertainty shows how in some cases it is not clear why a word is according to this method
  classified as anomalous
\item
  debiasing with the method provided by \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) does not solve the bias issues
\end{itemize}

One should remember that there is no baseline for the interpretation of the averaged cosine distances in (\protect\hyperlink{ref-Manzini2019blackToCriminalToCriminal}{\textbf{Manzini2019blackToCriminalToCriminal?}}). One may assume that if the cosine distance is close to 1 then it is a desired outcome as it means, according to cosine similarity assumptions, that there is almost no similarity between the words. However what does it mean to be close to 0? If the averaged cosine similarity is equal to 80, then should we still debiase it? It is unclear what the criteria are. On one hand, it seems to be beneficial when the outcome is simplified as it is easier to compare results with one value per set. On the other hand, it is prone to misunderstanding of how to interpret the results and what threshold to assume.

\hypertarget{what-we-propose-and-how-it-can-help}{%
\section{What we propose and how it can help:}\label{what-we-propose-and-how-it-can-help}}

\begin{itemize}
\tightlist
\item
\end{itemize}

@@

\leavevmode\hypertarget{ref-Manzini2019blackToCriminal}{}%
Manzini, T., Lim, Y. C., Tsvetkov, Y., \& Black, A. W. (2019). Black is to criminal as caucasian is to police: Detecting and removing multiclass bias in word embeddings. Retrieved from \url{http://arxiv.org/abs/1904.04047}
